---
title: "Model2_improvement"
author: "Katrien Quintelier"
date: "2023-06-09"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file()) # change wd from path dir to project dir for all chunks
getwd()
```

# Copy existing FlowSOM model
```{r}
dir.create("tmp_manualFlowSOM")
file.copy(from = "tmp/CytoNorm_FlowSOM.RDS", 
          to = "tmp_manualFlowSOM/CytoNorm_FlowSOM.RDS")
```

# Modify FlowSOM clustering
```{r}
fsom <- readRDS("tmp_manualFlowSOM/CytoNorm_FlowSOM.RDS")
fsom_tmp <- NewData(fsom = fsom,
                    input = read.FCS("tmp/before/agg_P1_C1.fcs"))
clustering <- GetClusters(fsom_tmp)

df <- data.frame("clustering" = clustering, 
                 "manual" = manual_agg[["agg_P1_C1"]])
counts <- table(df)

labels <- colnames(counts)[apply(counts,1,which.max)]

fsom <- UpdateMetaclusters(fsom = fsom, clusterAssignment = labels)

saveRDS(fsom, "tmp_manualFlowSOM/CytoNorm_FlowSOM.RDS")
```

# New model 1
```{r}
# Train model
manual_model1_withoutControl <- CytoNorm.train(files = flowSet(agg_P1_C1, agg_P2_C1),
                                               labels = c("P1", "P2"),
                                               outputDir = "tmp_manualFlowSOM",
                                               channels = cellTypeChannels,
                                               transformList = NULL,
                                               seed = 1,
                                               verbose = TRUE,
                                               plot = TRUE,
                                               FlowSOM.params = list(nCells = 1e+06, 
                                                                     xdim = 7, ydim = 7, 
                                                                     nClus = 3, 
                                                                     scale = FALSE))

# Normalize files
CytoNorm.normalize(model = manual_model1_withoutControl,
                   files = c(files_P1_C1, files_P2_C1),
                   labels = c(rep("P1", length(files_P1_C1)),
                              rep("P2", length(files_P2_C1))),
                   transformList = NULL,
                   verbose = TRUE,
                   prefix = "",
                   transformList.reverse = NULL, 
                   outputDir = "Data/Normalized_cellType_manualFlowSOM")
```

# New model 2
```{r}
# Extract goal distribution
goal_q <- CytoNorm:::getCytoNormQuantiles(manual_model1_withoutControl)

# Train model
manual_model2_towardsDistribution <- CytoNorm.train(files = c(files_P1_C2, files_P2_C2),
                                                    labels = c(rep(x = "P1",times = length(files_P1_C2)),
                                                               rep(x = "P2",times = length(files_P2_C2))),
                                                    outputDir = "tmp_manualFlowSOM",
                                                    channels = cellTypeChannels,
                                                    transformList = NULL,
                                                    seed = 1,
                                                    plot = TRUE,
                                                    verbose = TRUE,
                                                    normParams = list("goal" = goal_q))

# Normalize files
CytoNorm.normalize(model = manual_model2_towardsDistribution,
                   files = c(files_P1_C2, files_P2_C2),
                   labels = c(rep(x = "P1",times = length(files_P1_C2)),
                              rep(x = "P2",times = length(files_P2_C2))),
                   transformList = NULL,
                   verbose = TRUE,
                   prefix = "",
                   transformList.reverse = NULL, 
                   outputDir = "Data/Normalized_cellType_manualFlowSOM")
```

# Evaluation
## Make aggregates (same cells as previous aggregates)
```{r}
dir.create("tmp_manualFlowSOM/after")
aggregates <- list("agg_P1_C1" = files_P1_C1,
                   "agg_P1_C2" = files_P1_C2,
                   "agg_P2_C1" = files_P2_C1,
                   "agg_P2_C2" = files_P2_C2)

for (agg in names(aggregates)){
  writeLines(agg)
  files <- aggregates[[agg]]
  set.seed(2023)
  before <- read.FCS(paste0("tmp/before/", agg, ".fcs"))

  after <- before
  
  manual <- c()
  for (i_file in unique(before@exprs[,"File"])){
    ff <- read.FCS(paste0("Data/Normalized_cellType_manualFlowSOM/", basename(aggregates[[agg]][i_file])))
    i_IDs <- before@exprs[before@exprs[,"File"] == i_file,"Original_ID2"]
    after@exprs[before@exprs[,"File"] == i_file,1:(ncol(after@exprs)-3)] <- ff@exprs[i_IDs, ]
  }    
  write.FCS(after, paste0("tmp_manualFlowSOM/after/", agg, ".fcs"))
}
```

## Define cell types of interest
```{r}
cellTypes <- c("unlabeled",
               "All B cells",
               "Transitional B cells",
               "Naive B cells",
               "IgD+ memory B cells",
               "IgM+ memory B cells", 
               "Class switched memory B cells", 
               "Double negative B cells")

files_P1_C1_norm_manualFlowSOM <- list.files(path = "Data/Normalized_cellType_manualFlowSOM/", 
                                             pattern = "ID[1-4]_Panel1_TP..fcs", full.names = TRUE)
files_P2_C1_norm_manualFlowSOM <- list.files(path = "Data/Normalized_cellType_manualFlowSOM/", 
                                             pattern = "ID[1-4]_Panel2_TP..fcs", full.names = TRUE)
files_P1_C2_norm_manualFlowSOM <- list.files(path = "Data/Normalized_cellType_manualFlowSOM/", 
                                             pattern = "ID[5-8]_Panel1_TP..fcs", full.names = TRUE)
files_P2_C2_norm_manualFlowSOM <- list.files(path = "Data/Normalized_cellType_manualFlowSOM/", 
                                             pattern = "ID[5-8]_Panel2_TP..fcs", full.names = TRUE)

metrics <- list("Model1_before" = list("files" = c(files_P1_C1, "tmp/before/agg_P1_C1.fcs", 
                                                   "tmp/before/agg_P2_C1.fcs", files_P2_C1),
                                       "channels" = cellTypeChannels),
                "Model1_after" = list("files" = c(files_P1_C1_norm_manualFlowSOM, "tmp_manualFlowSOM/after/agg_P1_C1.fcs", 
                                                  "tmp_manualFlowSOM/after/agg_P2_C1.fcs", files_P2_C1_norm_manualFlowSOM),
                                       "channels" = cellTypeChannels),
                "Model2_before" = list("files" = c(files_P1_C1_norm_manualFlowSOM, files_P2_C1_norm_manualFlowSOM, "tmp_manualFlowSOM/after/agg_P1_C1.fcs", "tmp/after/agg_P2_C1.fcs",
                                                   "tmp/before/agg_P1_C2.fcs", "tmp/before/agg_P2_C2.fcs", files_P1_C2, files_P2_C2),
                                       "channels" = cellTypeChannels),
                "Model2_after" = list("files" = c(files_P1_C1_norm_manualFlowSOM, files_P2_C1_norm_manualFlowSOM, "tmp_manualFlowSOM/after/agg_P1_C1.fcs", "tmp_manualFlowSOM/after/agg_P2_C1.fcs",
                                                  "tmp_manualFlowSOM/after/agg_P1_C2.fcs", "tmp_manualFlowSOM/after/agg_P2_C2.fcs", files_P1_C2_norm_manualFlowSOM, files_P2_C2_norm_manualFlowSOM),
                                       "channels" = cellTypeChannels))
```

## Calculate EMD and MAD
```{r}
EMDs_manualFlowSOM <- list()
MADs_manualFlowSOM <- list()
for (model in names(metrics)){
  print(model)
  files <- metrics[[model]][["files"]]
  channels <- metrics[[model]][["channels"]]
  
  EMDs_manualFlowSOM[[model]] <- emdEvaluation(files = files,
                                 channels = channels,
                                 manual = manualList, return_all = TRUE, manualThreshold = 20)
  MADs_manualFlowSOM[[model]] <- CytoNorm:::madEvaluation(files = files[!startsWith(files, "tmp")], 
                                 channels = channels, return_all = TRUE,
                                 manual = manualList, transformList = NULL)
}
```

### Third approach: based on aggregates
```{r}
EMD_scores_manualFlowSOM <- list()
EMD_scores_manualFlowSOM[["aggregates"]] <- list()

# M1
for (subset in names(EMDs_manualFlowSOM)[1:2]){
  dist <- EMDs_manualFlowSOM[[subset]]$distances
  EMD_scores_manualFlowSOM[["aggregates"]][[subset]] <- matrix(NA,
                                                 nrow = length(dist), ncol = length(dist[[1]]),
                                                 dimnames = list(names(dist),
                                                                 names(dist[[1]])))
  for (cellType in names(dist)){
      for (marker in names(dist[[cellType]])){
        EMD_scores_manualFlowSOM[["aggregates"]][[subset]][cellType, marker] <- dist[[cellType]][[marker]][13, 14]
      }
  }
}

# M2
for (subset in names(EMDs_manualFlowSOM)[3:4]){
  dist <- EMDs_manualFlowSOM[[subset]]$distances
  EMD_scores_manualFlowSOM[["aggregates"]][[subset]] <- matrix(NA,
                                                 nrow = length(dist), ncol = length(dist[[1]]),
                                                 dimnames = list(names(dist),
                                                                 names(dist[[1]])))
  for (cellType in names(dist)){
      for (marker in names(dist[[cellType]])){
        EMD_scores_manualFlowSOM[["aggregates"]][[subset]][cellType, marker] <- max(dist[[cellType]][[marker]][25:26, 27:28], na.rm = T)
      }
  }
}
```

<!-- # Make figures -->
<!-- ```{r} -->
<!-- pdf("Results/Metrics_manualFlowSOM.pdf", width = 9, height = 4) -->
<!-- print(ggplot() + theme_void()) -->
<!-- for (subset in unique(sub("_.*", "", names(EMD_scores_manualFlowSOM[[1]])))){ -->
<!--   markerlevels <- NULL -->
<!--   plotlist <- list() -->
<!--   for (eval in names(EMD_scores_manualFlowSOM)){ -->
<!--     # EMD plot -->
<!--     EMD_before <- EMD_scores_manualFlowSOM[[eval]][[paste0(subset, "_before")]][2:length(cellTypes),] -->
<!--     EMD_after <- EMD_scores_manualFlowSOM[[eval]][[paste0(subset, "_after")]][2:length(cellTypes),] -->

<!--     EMD_df <- data.frame("before" = c(EMD_before), -->
<!--                          "after" = c(EMD_after), -->
<!--                          "feature" = paste(rownames(EMD_before), rep(colnames(EMD_before), each = nrow(EMD_before)), sep = "_")) -->
<!--     if (is.null(markerlevels)){ -->
<!--       markerlevels <- unique(unique(sub(".*_", "", EMD_df$feature))) -->
<!--     } -->
<!--     EMD_df$marker <- factor(sub(".*_", "", EMD_df$feature), levels = markerlevels) -->
<!--     EMD_df$celltype <- factor(sub("_.*", "", EMD_df$feature), levels = cellTypes) -->
<!--     max <- max(EMD_df[,1:2], na.rm = T) -->
<!--     EMD_df$worse <- EMD_df$before - EMD_df$after < 0 -->
<!--     EMD_df$feature <- ifelse(abs(EMD_df$before - EMD_df$after) < 10, "", EMD_df$feature) -->
<!--     p_emd <- ggplot(EMD_df, aes(x = after, y = before)) + -->
<!--       xlim(0,max) + ylim(0,max) + -->
<!-- #      geom_point(aes(color = worse)) + -->
<!--       geom_point(aes(color = marker, shape = celltype)) + -->
<!-- #      scale_color_manual(values = c("black", "red"))+ -->
<!--       geom_abline() + -->
<!--       scale_shape_manual(values = c(16, 17, 15, 3, 7, 8, 6)) + -->
<!-- #      ggrepel::geom_text_repel(aes(label = feature),# color = worse),  -->
<!-- #                               size = 2, max.overlaps = Inf) + -->
<!--       ggtitle(paste0(subset, " ", eval, " EMD")) + -->
<!--       theme_minimal() -->

<!--   leg_emd <- get_legend(p_emd) -->

<!--   p_emd <- p_emd  + -->
<!--       theme(legend.position = "none") -->
<!--   plotlist[[length(plotlist)+1]] <- p_emd     -->
<!--   } -->

<!--   names(markerlevels) <- sub(".*<(.*)>.*", "\\1", markerlevels) -->

<!--   # MAD plot -->
<!--   MAD_before <- MADs_manualFlowSOM[[paste0(subset, "_before")]][["comparison"]][-1,] -->
<!--   MAD_after <- MADs_manualFlowSOM[[paste0(subset, "_after")]][["comparison"]][-1,] -->
<!--   MAD_df <- data.frame("before" = c(MAD_before), -->
<!--                        "after" = c(MAD_after), -->
<!--                        "feature" = paste(rownames(MAD_before), rep(colnames(MAD_before), each = nrow(MAD_before)), sep = "_")) -->
<!--   MAD_df$marker <- factor(markerlevels[sub(".*_", "", MAD_df$feature)], levels = markerlevels) -->
<!--   MAD_df$celltype <- factor(sub("_.*", "", MAD_df$feature), levels = cellTypes) -->
<!--   max <- max(MAD_df[,1:2], na.rm = T) -->
<!--   MAD_df$bad <- abs(MAD_df$before - MAD_df$after) > 0.25 -->
<!--   MAD_df$feature <- ifelse(abs(MAD_df$before - MAD_df$after) < 0.25, "", MAD_df$feature) -->
<!--   p_mad <- ggplot(MAD_df, aes(x = after, y = before)) + -->
<!--     xlim(0,max) + ylim(0,max) + -->
<!-- #    geom_point(aes(color = bad)) + -->
<!--     geom_point(aes(color = marker, shape = celltype)) + -->
<!-- #    scale_color_manual(values = c("black", "red"))+ -->
<!--     geom_abline() + -->
<!--     scale_shape_manual(values = c(16, 17, 15, 3, 7, 8, 6)) + -->
<!-- #    ggrepel::geom_text_repel(aes(label = feature, color = bad),  -->
<!-- #                             size = 2, max.overlaps = Inf) + -->
<!--     ggtitle(paste0(subset, " MAD")) + -->
<!--     theme_minimal() -->

<!--   leg_mad <- get_legend(p_mad) -->

<!--   p_mad <- p_mad  + -->
<!--       theme(legend.position = "none") -->

<!--   plotlist[[length(plotlist)+1]] <- p_mad -->

<!--   print(ggarrange(ggarrange(plotlist = plotlist), -->
<!--                   as_ggplot(leg_emd), ncol=2, widths = c(4,1.5))) -->
<!-- } -->
<!-- dev.off() -->
<!-- ``` -->

<!-- # Heatmap plots -->
<!-- ```{r} -->
<!-- pdf("Results/distanceHeatmaps_manualFlowSOm.pdf") -->
<!-- print(ggplot() + theme_void()) -->
<!-- for (subset in unique(sub("_.*", "", names(EMDs_manualFlowSOM)))){ -->
<!--   gaps <- c(12, 13, 14) -->
<!--   if (subset == "Model2"){ -->
<!--     gaps <- c(12, 24, 25, 26, 27, 28, 40) -->
<!--   } -->
<!--   for (population in names(EMDs_manualFlowSOM[[paste0(subset, "_before")]][["distances"]])[2:length(names(EMDs[[paste0(subset, "_before")]][["distances"]]))]){ -->
<!--     for (marker in names(EMDs_manualFlowSOM[[paste0(subset, "_before")]][["distances"]][[population]])){ -->
<!--       for (condition in c("_before", "_after")){ -->
<!--         tmp_m <- EMDs_manualFlowSOM[[paste0(subset, condition)]][["distances"]][[population]][[marker]] -->
<!--         colnames(tmp_m) <- rownames(tmp_m) <- basename(colnames(tmp_m)) -->
<!--         p <- pheatmap::pheatmap(mat = tmp_m, display_numbers = TRUE,  -->
<!--                                 cluster_rows = FALSE, cluster_cols = FALSE,  -->
<!--                                 breaks = seq(0,10,length.out=100), fontsize = ifelse(subset == "Model2", 5, 7), -->
<!--                                 gaps_row = gaps, gaps_col = gaps, -->
<!--                                 main = paste0(subset, condition, "\n ", population, "_", marker)) -->
<!--         print(p) -->
<!--       } -->
<!--     } -->
<!--   } -->
<!-- } -->
<!-- dev.off() -->
<!-- ``` -->






# Supplementary figure 3: Results plot for manual model 2
## Density plots
```{r}
p1_plotlist <- list()
leg <- NULL
for (TP in c("before", "after")){
  # Read in aggregates
  agg_P1_C1 <- read.FCS(paste0("tmp/after/agg_P1_C1.fcs"))
  agg_P1_C2 <- read.FCS(paste0("tmp/", TP, "/agg_P1_C2.fcs"))
  agg_P2_C1 <- read.FCS(paste0("tmp/after/agg_P2_C1.fcs"))
  agg_P2_C2 <- read.FCS(paste0("tmp/", TP, "/agg_P2_C2.fcs"))
  
  if (TP == "after"){
  agg_P1_C2 <- read.FCS(paste0("tmp_manualFlowSOM/", TP, "/agg_P1_C2.fcs"))
  agg_P2_C2 <- read.FCS(paste0("tmp_manualFlowSOM/", TP, "/agg_P2_C2.fcs"))
  }
  # Colnames
  df_agg_P1_C1 <- data.frame(exprs(agg_P1_C1), check.names = FALSE)
  colnames(df_agg_P1_C1) <- GetMarkers(agg_P1_C1, channels = colnames(df_agg_P1_C1))
  df_agg_P1_C1$Manual <- manual_agg[["agg_P1_C1"]]
  
  df_agg_P1_C2 <- data.frame(exprs(agg_P1_C2), check.names = FALSE)
  colnames(df_agg_P1_C2) <- GetMarkers(agg_P1_C2, channels = colnames(df_agg_P1_C2))
  df_agg_P1_C2$Manual <- manual_agg[["agg_P1_C2"]]
  
  df_agg_P2_C1 <- data.frame(exprs(agg_P2_C1), check.names = FALSE)
  colnames(df_agg_P2_C1) <- GetMarkers(agg_P2_C1, channels = colnames(df_agg_P2_C1))
  df_agg_P2_C1$Manual <- manual_agg[["agg_P2_C1"]]
  
  df_agg_P2_C2 <- data.frame(exprs(agg_P2_C2), check.names = FALSE)
  colnames(df_agg_P2_C2) <- GetMarkers(agg_P2_C2, channels = colnames(df_agg_P2_C2))
  colnames(df_agg_P2_C2)[which(colnames(df_agg_P2_C2) == "CD268 (BAFFR)")] <- "CD268"
  df_agg_P2_C2$Manual <- manual_agg[["agg_P2_C2"]]
  
  # Make data frame
  markers <- c("IgM", "CD27")
  names <-unique(c(paste0(cellTypeMarkers, " <", cellTypeChannels, ">"), 
                   paste0(P1_cellStateMarkers, " <", P1_cellStateChannels, ">"), 
                   paste0(P2_cellStateMarkers, " <", P2_cellStateChannels, ">")))
  
  df <- dplyr::bind_rows(df_agg_P1_C1,
                         df_agg_P1_C2,
                         df_agg_P2_C1,
                         df_agg_P2_C2)
  df$Batch <- c(rep(x = "P1_C1", times = nrow(agg_P1_C1)),
                rep(x = "P1_C2", times = nrow(agg_P1_C2)),
                rep(x = "P2_C1", times = nrow(agg_P2_C1)),
                rep(x = "P2_C2", times = nrow(agg_P2_C2)))
  df$File[df$Batch == "P1_C2"] <- df$File[df$Batch == "P1_C2"] + 12
  df$File[df$Batch == "P2_C1"] <- df$File[df$Batch == "P2_C1"] + 24
  df$File[df$Batch == "P2_C2"] <- df$File[df$Batch == "P2_C2"] + 36
  df$Manual <- factor(df$Manual, levels = cellTypes)
  df <- df[,c(markers, "Manual", "File", "Batch")]

  # Plot
  for (marker in markers){
  plotlist <- list()
    p <- ggplot(na.omit(df[,c(marker, "File", "Batch")]), aes(x=!!sym(marker), color = Batch)) +
      stat_density(aes(group = File), geom = "line", position = "identity", alpha = 0.2)+
      stat_density(geom = "line", position = "identity", linewidth = 0.7) +
      scale_color_manual(values = batch_colors)+
      xlim(c(-1,5)) +
      xlab(paste0(marker, " <", GetChannels(agg_P1_C1, marker), ">")) +
      theme_minimal() +
      ylab(ifelse(TP == "before", "original", "normalized"))#+
      #theme(legend.direction = "horizontal") 
    if (is.null(leg)){
      leg <- get_legend(p)
    }
    if (TP == "before" & marker == markers[1]){
      p <- p + ggtitle(label = "", subtitle = "All B cells")
    } 
    x_range <- layer_scales(p)$x$range$range
    p <- p + theme(legend.position = "none") +
       theme(axis.title.x=element_text(colour=ifelse(marker == "CD27", "#F8766D", "#C77CFF"))) 
    plotlist[[length(plotlist) + 1]] <- p
    
    for (cellType in cellTypes[-c(1,2)]){
      df_sub <- na.omit(df[df$Manual == cellType,c(marker, "File", "Batch")])
      p <- ggplot(df_sub, aes(x=!!sym(marker), color = Batch)) +
        stat_density(aes(group = File), geom = "line", position = "identity", alpha = 0.2)+
        stat_density(geom = "line", position = "identity", linewidth = 0.7) +
        scale_color_manual(values = batch_colors) +
        xlab(paste0(marker, " <", GetChannels(agg_P1_C1, marker), ">")) +
        theme_minimal() + 
        theme(legend.position = "none") +
        xlim(c(-1,5)) +
        ylab("") +
       theme(axis.title.x=element_text(colour=ifelse(marker == "CD27", "#F8766D", "#C77CFF"))) 
    if (TP == "before" & marker == markers[1]){
      p <- p + ggtitle(label = "", subtitle = cellType)  +
        theme(plot.subtitle = element_text(hjust = 0.5))
      }
      plotlist[[length(plotlist) + 1]] <- p
  
    }
  p1_plotlist[[length(p1_plotlist)+1]] <- ggarrange(plotlist = plotlist, ncol = 7, nrow = 1)
  }
}

p1_plotlist[[length(p1_plotlist)+1]] <- NULL

p1 <- ggarrange(ggarrange(plotlist = c(p1_plotlist[1], p1_plotlist[3], p1_plotlist[5], p1_plotlist[2], p1_plotlist[4]), ncol = 1, nrow = 5, heights = c(2.6, 2, 0.3, 2, 2)), 
                as_ggplot(leg), ncol = 2, nrow = 1, widths = c(10,1))
```

## EMD and MAD plot
```{r}
for (subset in "Model2"){
  markerlevels <- NULL
  plotlist <- list()
  for (eval in names(EMD_scores_manualFlowSOM)[1]){
    # EMD plot
    EMD_before <- EMD_scores_manualFlowSOM[[eval]][[paste0(subset, "_before")]][-2,]
    EMD_after <- EMD_scores_manualFlowSOM[[eval]][[paste0(subset, "_after")]][-2,]
    rownames(EMD_before)[1] <- "All B cells"
    rownames(EMD_after)[1] <- "All B cells"
    
    EMD_df <- data.frame("before" = c(EMD_before),
                         "after" = c(EMD_after),
                         "feature" = paste(rownames(EMD_before), rep(colnames(EMD_before), each = nrow(EMD_before)), sep = "_"))
    if (is.null(markerlevels)){
      markerlevels <- unique(unique(sub(".*_", "", EMD_df$feature)))
    }
    EMD_df$marker <- factor(sub(".*_", "", EMD_df$feature), levels = markerlevels)
    EMD_df$celltype <- factor(sub("_.*", "", EMD_df$feature), levels = cellTypes)
    max <- max(EMD_df[,1:2], na.rm = T)
    EMD_df$worse <- EMD_df$before - EMD_df$after < 0
    EMD_df$feature <- ifelse(abs(EMD_df$before - EMD_df$after) < 10, "", EMD_df$feature)
    p_emd <- ggplot(EMD_df, aes(x = after, y = before)) +
      xlim(0,max) + ylim(0,max) +
#      geom_point(aes(color = worse)) +
      geom_point(aes(color = marker, shape = celltype), size = 3) +#, size = 0.6)) +
#      scale_color_manual(values = c("black", "red"))+
      geom_abline() +
      scale_shape_manual(values = c(16, 17, 15, 3, 7, 8, 6)) +
#      ggrepel::geom_text_repel(aes(label = feature),# color = worse), 
#                               size = 2, max.overlaps = Inf) +
      ggtitle(paste0("(i) ", subset, " ", eval, " EMD")) +
      theme_minimal() + 
      labs(color = "Marker",
           shape = "Cell type") +
      theme(legend.position = "bottom", 
            legend.box="vertical")#, 
#            legend.margin=margin(),
#            legend.key.size = unit(1, 'cm'), 
#        legend.title = element_text(size=16), 
#        legend.text = element_text(size=12)) #+ 
#      guides(size = FALSE, colour = guide_legend(override.aes = list(size=4)),
#             shape = guide_legend(override.aes = list(size=4)))
  leg_emd <- get_legend(p_emd)

  p_emd <- p_emd  +
      theme(legend.position = "none") +
      xlab("normalized") +
      ylab("original")
  plotlist[[length(plotlist)+1]] <- p_emd    
  }
  
  names(markerlevels) <- sub(".*<(.*)>.*", "\\1", markerlevels)
  
  # MAD plot
  MAD_before <- MADs_manualFlowSOM[[paste0(subset, "_before")]][["comparison"]][-2,]
  MAD_after <- MADs_manualFlowSOM[[paste0(subset, "_after")]][["comparison"]][-2,]
  rownames(MAD_before)[1] <- "All B cells"
  rownames(MAD_after)[1] <- "All B cells"

  MAD_df <- data.frame("before" = c(MAD_before),
                       "after" = c(MAD_after),
                       "feature" = paste(rownames(MAD_before), rep(colnames(MAD_before), each = nrow(MAD_before)), sep = "_"))
  MAD_df$marker <- factor(markerlevels[sub(".*_", "", MAD_df$feature)], levels = markerlevels)
  MAD_df$celltype <- factor(sub("_.*", "", MAD_df$feature), levels = cellTypes)
  max <- max(MAD_df[,1:2], na.rm = T)
  MAD_df$bad <- abs(MAD_df$before - MAD_df$after) > 0.25
  MAD_df$feature <- ifelse(abs(MAD_df$before - MAD_df$after) < 0.25, "", MAD_df$feature)
  p_mad <- ggplot(MAD_df, aes(x = after, y = before)) +
    xlim(0,max) + ylim(0,max) +
#    geom_point(aes(color = bad)) +
    geom_point(aes(color = marker, shape = celltype), size = 3) +#, size = 0.6)) +
#    scale_color_manual(values = c("black", "red"))+
    geom_abline() +
    scale_shape_manual(values = c(16, 17, 15, 3, 7, 8, 6)) +
#    ggrepel::geom_text_repel(aes(label = feature, color = bad), 
#                             size = 2, max.overlaps = Inf) +
    ggtitle(paste0("(ii) ", subset, " MAD")) +
    theme_minimal()

  leg_mad <- get_legend(p_mad)

  p_mad <- p_mad  +
      theme(legend.position = "none") +
      xlab("normalized") +
      ylab("original")

  plotlist[[length(plotlist)+1]] <- p_mad
}

p2 <- plotlist[[1]]
p3 <- as.ggplot(leg_emd)
p4 <- plotlist[[2]]
```

## Aggregates contribution plot: barplot with cell type composition per aggregate file
```{r}
agg_name <- c()
man_labels <- c()
clustering <- c()

agg_objects <- c(agg_P1_C1, agg_P2_C1, agg_P1_C2, agg_P2_C2)
agg_names <- c("agg_P1_C1", "agg_P2_C1", "agg_P1_C2", "agg_P2_C2")
for (i in 1:length(agg_objects)){
  agg_name <- c(agg_name, rep(agg_names[i], nrow(agg_objects[[i]])))
  man_labels <- c(man_labels, as.character(manual_agg[[agg_names[i]]]))
  fsom_tmp <- NewData(manual_model1_withoutControl$fsom, agg_objects[[i]])
  clustering <- c(clustering, paste0("Cluster ", GetMetaclusters(fsom_tmp)))
}

df <- data.frame("Batch aggregate" = agg_name,
                 "Cell type" = man_labels,
                 "FlowSOM cluster" = clustering,
                 check.names = FALSE)

df <- data.frame(data.table::setDT(df)[,list(Contribution=.N),names(df)], check.names = FALSE)

df$`Batch aggregate` <- factor(df$`Batch aggregate`, 
                               levels = c("agg_P1_C1", "agg_P2_C1", "agg_P1_C2", "agg_P2_C2"))
df$`Cell type` <- factor(df$`Cell type`,
                         levels = c("Transitional B cells", "Naive B cells", 
                                    "IgD+ memory B cells", "IgM+ memory B cells", 
                                    "Class switched memory B cells", 
                                    "Double negative B cells", "unlabeled"))
df$`FlowSOM cluster` <- factor(df$`FlowSOM cluster`,
                               levels = paste0("Cluster ", 
                                               c("Transitional B cells", "Naive B cells", 
                                                 "IgD+ memory B cells", "IgM+ memory B cells", 
                                                 "Class switched memory B cells", 
                                                 "Double negative B cells", "unlabeled")))
levels(df$`FlowSOM cluster`) <- c("Cluster \n Transitional \n B cells", 
                                  "Cluster \n Naive \n B cells", 
                                  "Cluster \n IgD+ memory \n B cells", 
                                  "Cluster \n IgM+ memory \n B cells", 
                                  "Cluster \n Class switched \n memory B cells", 
                                  "Cluster \n Double negative \n B cells", 
                                  "Cluster \n unlabeled")

plotlist <- list()
i <- 1

# leg <- get_legend(ggplot(data = df, 
#                          aes(x = `Batch aggregate`, y = Contribution, fill = `Cell type`)) +
#                     geom_bar(position="fill", stat="identity") +
#                     scale_fill_manual(values = c("#fbf8cc", "#ffcfd2", "#b9fbc0", "#a3c4f3", "#cfbaf0", "#8eecf5", "#edede9")) +
#                     facet_wrap(~`FlowSOM cluster`,nrow=1))

#for (cluster in levels(df$`FlowSOM cluster`)){
  plotlist[[length(plotlist)+1]] <-  ggplot(data = df, aes(x = `Batch aggregate`, y = Contribution, fill = `Cell type`)) +
  geom_bar(position="fill", stat="identity") +
  scale_fill_manual(values = c("#fbf8cc", "#ffcfd2", "#b9fbc0", "#a3c4f3", "#cfbaf0", "#8eecf5", "#edede9")) +
  facet_wrap(~`FlowSOM cluster`,nrow=1) +
  xlab("") + ylab("Composition") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.2,
                                   colour = c("#900002", "#0046A1", 
                                              "#FF3437", "#6FCFFF")),
        axis.ticks = element_blank())+ 
    theme(strip.text.x = element_text(size = 8))#+
   # ggtitle(label = "", subtitle = cluster)  #+
    #theme(legend.title = element_text(size=16), 
    #      legend.text = element_text(size=12))

  # if (cluster != "Cluster Transitional B cells"){
  #   plotlist[[length(plotlist)]] <- plotlist[[length(plotlist)]] +
  #     theme(axis.title.y=element_blank(),
  #       axis.text.y=element_blank(),
  #       axis.ticks.y=element_blank())
  # }
  # plotlist[[length(plotlist)]] <- plotlist[[length(plotlist)]] +
  #     theme(legend.position = "none")
  
  i <- i + 1
#}
#plotlist[[length(plotlist)+1]] <- as_ggplot(leg)
#p5 <- ggarrange(plotlist = plotlist, nrow = 1, widths = c(2.85, 2, 2, 2, 2, 2, 4.4))
  p5 <- plotlist[[length(plotlist)]]
```


## Bring everything together
```{r}
pdf("Results/SupplementaryFigure4.pdf", width = 16, height = 14)
p <- ggarrange(ggarrange(ggarrange(
                           ggarrange(p2,p4, 
                                     nrow = 1, labels = c("A", "")),p3, NULL, 
                           ncol = 1, heights = c(3, 0.5, 0.1)), p5, 
                         nrow = 1, labels = c("", "B"), widths = c(2,2)), ggarrange(NULL, p1, nrow = 1, widths = c(0.5,40)), 
               ncol = 1, labels = c("", "C"), heights = c(3.5,5))
print(p)
dev.off()
```